---
title: "Creating publication-ready tables"
output:
  rmarkdown::html_vignette:
    number_sections: true
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{4. Creating publication-ready tables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates tools for creating publication-ready tables from both experimental and meta-analytic data. BLPlabtools provides two main categories of table creation functions:

1. **Meta-analysis moderator tables**: Functions for creating standardized tables showing effect sizes across moderator levels
2. **LaTeX table formatting utilities**: Functions for polishing tables created with `stargazer` or `knitr::kable()`

We'll cover both types of tables using realistic examples from the built-in datasets.

# Meta-analysis moderator tables

When conducting meta-analyses, you often need to create tables showing how effect sizes vary across moderator variables (study design, intervention type, population characteristics, etc.). BLPlabtools provides a suite of functions to automate this process:

- `process_group()`: Create complete moderator tables with one function (high-level)
- `extract_model_results()`: Extract overall meta-analysis results
- `run_subset_meta_analysis()`: Run meta-analysis on data subsets (for custom filtering)

**Note:** `process_group()` internally uses `run_meta_regression()` to generate p-values comparing moderator levels. You rarely need to call `run_meta_regression()` directly.

## Example data: Sexual violence interventions

We'll use the built-in `sv_data` dataset, which contains effect sizes from interventions to prevent sexual violence:

```{r load_data}
library(BLPlabtools)
library(dplyr)
library(knitr)

# Preview the data
sv_data |>
  select(unique_study_id, author, year, study_design, behavior_type, d, var_d) |>
  head()
```

This dataset includes:
- **73 effect sizes** from **29 unique studies**
- Moderators: `study_design`, `behavior_type`, `population`
- Effect sizes: Cohen's d (`d`) and variance (`var_d`)

## Overall meta-analysis results

Start by extracting overall results across all studies:

```{r overall_results}
# Overall effect across all studies
overall <- sv_data |> extract_model_results(approach_name = "Overall")
kable(overall)
```

The `extract_model_results()` function runs `metafor::rma()` with cluster-robust standard errors and extracts key statistics in a clean format: number of studies and estimates, effect size (Delta), standard error, confidence interval, p-value, and tau (residual heterogeneity).

## Creating moderator tables with process_group()

The `process_group()` function creates complete moderator analysis tables with one call. It runs separate meta-analyses for each level of a moderator and adds p-values comparing each level to a reference:

```{r study_design_table}
# Create moderator table for study design
design_table <- sv_data |>
  process_group(
    group_var = "study_design",
    ref_level = "Randomized Controlled Trial",
    include_tau = TRUE
  )

kable(design_table)
```

The output includes:
- `Moderator`: The level name (Randomized Controlled Trial, Quasi-Experimental, etc.)
- `N_Studies` and `N_Estimates`: Sample sizes
- `Delta`: Effect size estimate for this level
- `tau`: Residual heterogeneity
- `CI`: Confidence interval
- `p_val`: P-value testing if this level's effect differs from zero
- `p_val_ref`: P-value comparing this level to the reference level (Randomized Controlled Trial)

Notice "ref" for Randomized Controlled Trial's `p_val_ref`—it's the reference level, so other levels are compared to it.

## Custom level ordering

Control the order of moderator levels in the output:

```{r custom_order}
# Custom ordering for behavior type
behavior_table <- sv_data |>
  process_group(
    group_var = "behavior_type",
    ref_level = "Ideas",
    order_levels = c("Ideas", "Perpetration", "Victimization", "Bystander", "Involvement"),
    include_tau = FALSE
  )

kable(behavior_table)
```

## Manual subset analysis

For more control, use `run_subset_meta_analysis()` directly:

```{r manual_subset}
# Analyze just RCTs
rct_only <- sv_data |>
  run_subset_meta_analysis(
    group_var = "study_design",
    level = "Randomized Controlled Trial",
    approach_name = "RCTs only",
    include_tau = TRUE
  )

kable(rct_only)
```

You can also filter by string patterns:

```{r string_filter}
# Analyze studies with "college" in population
college_studies <- sv_data |>
  run_subset_meta_analysis(
    filter_column = "population",
    filter_string = "college",
    str_detect_flag = TRUE,
    approach_name = "College students"
  )

kable(college_studies)
```

## Combining multiple approaches

Create publication-ready tables comparing different analytical approaches:

```{r combined_table}
# Combine overall + moderator analyses
full_table <- bind_rows(
  sv_data |> extract_model_results(approach_name = "Overall"),
  sv_data |>
    filter(study_design == "RCT") |>
    extract_model_results(approach_name = "RCTs only"),
  sv_data |>
    filter(behavior_type == "Perpetration") |>
    extract_model_results(approach_name = "Perpetration outcomes")
)

kable(full_table)
```

# Experimental data table formatting

For experimental data, BLPlabtools provides utilities to format LaTeX tables created with `stargazer` or `knitr::kable()`. These functions work by manipulating captured LaTeX output as text.

## Basic number and text formatting

Format numbers consistently across tables:

```{r number_formatting}
# Format to 2 decimal places
values <- c(0.1543, 2.567, 10.2)
two_digits(values)

# Add parentheses (for standard errors)
add_parentheses(two_digits(values))

# Add brackets (for confidence intervals)
add_brackets(two_digits(values))
```

**Other formatting functions:**
- `word_italicize()`: Italicize specific words in LaTeX output

## Working with data frames

Utilities for adding rows to tables:

```{r dataframe_utils, eval=FALSE}
# Add header row
my_table |> add_top_row(c("Model 1", "Model 2", "Model 3"))

# Add footer row
my_table |> add_bottom_row(c("Note: All models include controls"))

# Placeholder for pipe chains
my_table |>
  some_operation() |>
  nothing()  # Does nothing, useful at end of chains
```

## LaTeX table utilities

Functions for working with captured LaTeX output:

```{r latex_utils, eval=FALSE}
# Capture table output
my_table <- capture.output(kable(head(sv_data), format = "latex"))

# Display cleanly in console
print_table(my_table)

# Add cross-reference label
my_table |> table_label(caption = "sv_results")

# Customize table numbering
my_table |> table_numbers(table_number = "2a")

# Resize table
my_table |> kable_resize(size_in_inches = 5)
```

## Stargazer-specific formatting

Customize `stargazer` output:

```{r stargazer_format, eval=FALSE}
library(stargazer)

# Create example regression
model <- lm(d ~ as.numeric(study_design == "RCT"), data = sv_data)

# Capture stargazer output
table_output <- capture.output(
  stargazer(model, type = "latex", report = "csp")
)

# Change p-value format to brackets
table_output |> stargazer_pvalues("brackets")

# Add custom footnote
table_output |> add_endnote(
  note = "Standard errors clustered by study. Controls included.",
  size_in_inches = 6,
  rm.stargazer.stars = TRUE
)

# Customize row names
table_output |> stargazer_rowname(old_name = "RCT", new_name = "Treatment (RCT)")
```

# Complete workflow example

Here's a typical workflow for creating publication tables from meta-analytic data:

```{r complete_workflow, eval=FALSE}
# 1. Create moderator table
moderator_results <- sv_data |>
  process_group(
    group_var = "study_design",
    ref_level = "Randomized Controlled Trial",
    include_tau = TRUE
  )

# 2. Format for LaTeX
library(knitr)
latex_table <- kable(
  moderator_results,
  format = "latex",
  caption = "Effect sizes by study design",
  booktabs = TRUE
) |>
  capture.output()

# 3. Add formatting
formatted_table <- latex_table |>
  table_label(caption = "moderator_analysis") |>
  print_table()

# 4. Save to file
writeLines(formatted_table, "tables/moderator_table.tex")
```

# Key takeaways

**For meta-analysis tables:**
- Use `process_group()` for complete moderator tables with one call
- Use `extract_model_results()` for overall results or combining approaches
- Use `run_subset_meta_analysis()` for custom filtering and granular control
- Always specify a reference level for meta-regression comparisons

**For table formatting:**
- Capture LaTeX output with `capture.output()`
- Use `print_table()` to preview in console
- Combine formatting functions with pipes
- Add `table_label()` for cross-referencing in LaTeX documents

**Best practices:**
- Use consistent number formatting (e.g., `two_digits`) across all tables
- Include descriptive notes explaining moderator coding and analytical choices
- Test table rendering in your LaTeX document early
- Keep tau in moderator tables when between-study heterogeneity is important

For complete function documentation, see:
- Meta-analysis: `?process_group`, `?run_subset_meta_analysis`, `?extract_model_results`
- Formatting: `?table_prep`, `?star_ready`
